The Effects of weather on maize yields: New evidence from Kenya

### 2018-12-24 writing/draft3/draft17.tex
writing
correcting

\pdfminorversion=4
\submitto{\ERL}
\makesavenoteenv{tabular}

@@

In this study, we analyzed the relationship between maize yields and climate. Our dataset consists of an yearly panel of $47$ counties of Kenya describing the period of $1981-2017$. 
	We acquired the county level yearly yield data from the Famine Early Warning Systems Network (FEWS NET). Regarding the weather data, we exploited $0.25^\circ$ resolution precipitation and temperature gridded datasets. The precipitation data were obtained from the Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS) while the temperature data are available at the website of the Berkeley Earth. We calculated spatial averages of the gridded data over the counties to get a single value for each county and each point in time. The frequency of precipitation and temperature data was daily, therefore, further aggregation was needed to obtain yearly values corresponding to the yearly frequency of the yield data. Hence, each weather characteristic was further aggregated over years resulting in a county-level panel conformable with the yield data.

	There are two predominant rainfall regimes in Kenya. The arid and semi-arid (ASAL) counties exhibit mostly bi-modal precipitation patterns with long rains lasting from March to May and short rains occurring between October and December. In the non-ASAL counties, the single rainy season usually starts in March and lasts until August. Following the precipitation patterns and closely related planting and harvesting calendar, we computed yearly values of various weather characteristics as follows: For the ASAL counties, we used daily data covering October, November and December of the previous year and March, April, May of the current year and for the non-ASAL counties we used daily data covering March to August of the current year.

Following the procedure described above, we calculated a number of previously used characteristics of precipitation and temperature including indicators of floods (for the list of these measures see table~\ref{vars} in the appendix). The measures which we found significant were: Total seasonal precipitation and its squared values, average seasonal temperature, coefficient of variation of seasonal precipitation, coefficient of variation of average seasonal temperature (converted into Kelvin because other scales of temperature such us Celsius are interval and coefficient of variation does not have any meaning for data measured on an interval scale),  maximum length of dry spell in number of days and the number of dry spells lasting for four days or more (we considered every day when precipitation didn't exceed $1$ mm as a dry day). We did not include squared values of the average seasonal temperature as it was not significant in our models and there is not much evidence of non-linear relationship between yield and temperature in the past literature.
\subsection{Statistical approach}\label{stats}

@@

where $\mathbf{X}$ is an $n \times p$ model matrix of fixed effects, $\mathbf{\beta}$ is a $p$-dimensional fixed-effects parameter, $\mathbf{Z}$ is an $n \times q$ model matrix for the $q$-dimensional vector of random-effects variable $\mathscr{B}$ evaluated at $\mathbf{b}$ and $\sigma$ a scale factor. The distribution of $\mathscr{B}$ can be written as:

@@

It was apparent from the histogram of the maize yield data that the distribution of maize yield was closer to a log-normal than to a normal distribution. Therefore, we opted for a log-linear functional form of our models as is a common practice in agricultural economics. Furthermore, we scaled all predictors by subtracting mean and dividing them by standard deviation to avoid convergence problems.

To find the most suitable set of fixed effects, we adapted a stepwise selection procedure which minimized the Akaike's Information Criterion (AIC). The first step of the procedure was the estimation of the model with the complete set of our weather measures in the fixed effects and with the random intercepts. After this, we performed a search through the subsets of the fixed effects minimizing the AIC and allowing for steps in both `backward' and `forward' directions (that is removing and adding predictors). We decided not to include number of heatwave days and average maximum daily temperature although including them as fixed effects would slightly improve the AIC. The reasons for not including them were their strong correlation with average seasonal temperature (the correlation coefficient was $0.981$ in case of maximum daily temperature and $0.517$ for the number of heatwave days with the p-values smaller than $1\times10^{-6}$ in both cases), high variance inflation factor (VIF) of the maximum temperature ($5.397$) and the lack of significance of their individual t-tests (the p-values were $0.121$ for the maximum daily temperature and $0.075$ for the number of heatwave days). Furthermore, including the heatwave days and maximum daily temperature would only lead to small change in AIC from $2080.877$ to $2076.439$ (these values differ from the AIC presented in table~\ref{ARMA} in the appendix because the final estimates reported in table~\ref{ARMA} were estimated using the restricted maximum likelihood (REML) whereas the models were re-estimated using the ML method for comparison\footnote{In general, REML is the preferred estimation method for the mixed effects models as the ML estimates of the variance component are biased, however, REML criterion is not meaningful for comparison of the models with different fixed effects. Therefore, the models had to be re-estimated by ML for comparison \cite{Zuur2009}.}). In addition, the coefficients of both heatwave days and maximum daily temperature were positive, which is opposite to what we expected. Hence, we concluded that the decrease in AIC and the positive signs are likely to be results of the correlation with average temperature and we did not include them in our preferred model.

We further applied a series of likelihood ratio tests to test for significance of random slopes. No subset of the random slopes improved the fit, therefore besides the fixed effects, our final model only included the random intercepts.\footnote{Although adding the random slope of the coefficient of variation slightly improved the AIC from $2122.226$ to $2122.110$, we decided to not to include the coefficient of variation of temperature in the random effects. The reason for not including the random slope is that the difference in AIC is negligible and the F-test of the comparison of the two models is insignificant favouring the simpler model (p-value$=0.128$).}

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. We started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of the procedure, the insignificant random effects were removed one by one. In each iteration, the insignificant variable with the highest p-value was removed . This was repeated until only the significant predictors remained. After this, we applied an analogous procedure to the fixed effects until we were left only with the significant ones. As recommended in literature, we considered the level of significance $\alpha=0.1$ for random effects and $\alpha=0.05$~ for fixed effects \cite{lmerTest}. This method led to the same model as the one obtained with our primary AIC-based method described above.

>>>>

% change the following at uni. 24.12.2018
%\pdfminorversion=4
\submitto{\ERL}
\makesavenoteenv{tabular}

@@

In this study, we analysed the relationship between maize yields and climate. Our dataset consisted of an yearly panel of $47$ counties of Kenya describing the period of $1981-2017$. 
	We acquired the county level yearly yield data from the Famine Early Warning Systems Network (FEWS NET). Regarding the weather data, we exploited $0.25^\circ$ resolution precipitation and temperature gridded datasets. The precipitation data were obtained from the Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS) while the temperature data are available at the website of the Berkeley Earth. We calculated spatial averages of the gridded data over the counties to get a single value for each county and each point in time. The frequency of precipitation and temperature data was daily, therefore, further aggregation was needed to obtain yearly values corresponding to the yearly frequency of the yield data. Hence, each weather characteristic was further aggregated over years resulting in a county-level panel conformable with the yield data.

	There are two predominant rainfall regimes in Kenya. The arid and semi-arid (ASAL) counties exhibit mostly bi-modal precipitation patterns with long rains lasting from March to May and short rains occurring between October and December. In the non-ASAL counties, the single rainy season usually starts in March and lasts until August. Following the precipitation patterns and the closely related planting and harvesting calendar, we computed yearly values of various weather characteristics as follows: For the ASAL counties, we used daily data covering October, November and December of the previous year and March, April, May of the current year and for the non-ASAL counties we used daily data covering March to August of the current year.

Following the procedure described above, we calculated a number of previously used characteristics of precipitation and temperature including indicators of floods. These measures are listed in table~\ref{vars} in the appendix. The measures which we found significant were: Total seasonal precipitation and its squared values, average seasonal temperature, coefficient of variation of seasonal precipitation, coefficient of variation of average seasonal temperature (converted into Kelvin because the Celsius scale is interval and the coefficient of variation does not have any meaning for data measured on an interval scale),  maximum length of dry spell measured as number of consecutive dry days and the number of dry spells lasting for four days or more (for the purpose of this study we defined a dry day as a day when precipitation didn't exceed $1$mm). We did not include squared values of the average seasonal temperature in our preferred models as it was not significant and most of the past literature has not supported the evidence of the non-linear relationship between yield and temperature.
\subsection{Statistical approach}\label{stats}

@@

where $\mathbf{X}$ is an $n \times p$ model matrix of fixed effects, $\mathbf{\beta}$ is a $p$-dimensional fixed-effects parameter, $\mathbf{Z}$ is an $n \times q$ model matrix for the $q$-dimensional vector of random-effects variable $\mathscr{B}$ evaluated at $\mathbf{b}$ and $\sigma$ is a scale factor. The distribution of $\mathscr{B}$ can be written as:

@@

It was apparent from the histogram of the maize yield data that the distribution of maize yield was closer to a log-normal than to a normal distribution. Therefore, we opted for a log-linear functional form of our models as is a common practice in the agricultural economics. Furthermore, we scaled all predictors by subtracting mean and dividing them by standard deviation to avoid convergence problems.

To find the most suitable set of fixed effects, we adapted a stepwise selection procedure which minimized the Akaike's Information Criterion (AIC). The first step of the procedure was the estimation of the model with the complete set of our weather measures in the fixed effects and with the random intercepts. After this, we performed a search through the subsets of the fixed effects minimizing the AIC and allowing for steps in both `backward' and `forward' directions (that is removing and adding the predictors). We decided not to include number of heatwave days and average maximum daily temperature although including them as fixed effects would slightly improve the AIC. The reasons for not including them were their strong correlation with average seasonal temperature (the correlation coefficient was $0.981$ in the case of maximum daily temperature and $0.517$ for the number of heatwave days with the p-values smaller than $1\times10^{-6}$ in both cases), high variance inflation factor (VIF) of the maximum temperature ($5.397$) and the lack of significance of their individual t-tests (the p-values were $0.121$ for the maximum daily temperature and $0.075$ for the number of heatwave days). Furthermore, including heatwave days and maximum daily temperature would only lead to a small change in the AIC from $2080.877$ to $2076.439$ (these values differ from the AIC presented in table~\ref{ARMA} in the appendix because the final estimates reported in table~\ref{ARMA} were estimated using the restricted maximum likelihood (REML) whereas the models were re-estimated using the ML method for comparison\footnote{REML is generally the preferred estimation method for the mixed effects models as the ML estimates of the variance component are biased, however, the REML criterion is not meaningful for comparison of the models with different fixed effects. Therefore, we re-estimated the models by ML for the purpose of comparison \cite{Zuur2009}.}). In addition, the coefficients of both heatwave days and maximum daily temperature were positive, which is opposite to what we expected. Hence, we concluded that the decrease in the AIC and the positive signs are likely to be results of the high correlation with average temperature and we did not include them in our preferred specification.

We further applied a series of likelihood ratio tests to test for the significance of random slopes. No subset of the random slopes improved the fit; therefore, besides the fixed effects, our final model only included the random intercepts.\footnote{Although adding the random slope of the coefficient of variation improved the AIC from $2122.226$ to $2122.110$, we decided to not to include the coefficient of variation of temperature in the random effects. The reason for not including the random slope is that the difference in AIC is negligible and the F-test of the comparison of the two models is insignificant favouring the simpler model (p-value$=0.128$).}

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. We started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of the procedure, the insignificant random effects were removed one by one. In each iteration, the insignificant variable with the highest p-value was removed. This was repeated until only the significant predictors remained. After this, we applied an analogous procedure to the fixed effects until we were left only with the significant variables. As recommended in literature, we considered the level of significance $\alpha=0.1$ for random effects and $\alpha=0.05$~ for fixed effects \cite{lmerTest}. This method led to the same model as the one obtained with our primary AIC-based method described above.
